---
title: "Session Eleven: R Markdown"
author: "Estimated Duration: 40 minutes"
output:
  html_document: 
    code_folding: show
    highlight: tango
    number_sections: no
    theme: sandstone
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br>

[__<<__ Back to Menu](rtutorial_basic.html)  
[__>>__ Session Twelve](RTutorial_12_Flexdashboard-htmlwidgets-Shiny.html)

# {.tabset}

## R Markdown File

### __R Markdown File__

<br>

#### Create RMarkdown File

In RStudio, go to File > New File > RMarkdown  

Work on the template that comes with the new file.  
Refer to RMarkdown __cheatsheet__.

<br>

## Tables

### __Tables__

<br>

#### Standard RMarkdown Table

Example below displays table with `df_print: paged` defined in YAML  
Other df_print options are `default`, `kable` and `tibbl`  

```{r}

df1 <- data.frame(x=letters[1:5], y=rnorm(5), z=runif(5)>0.5, status=factor(c("Open","Closed","Closed","Open","Pending")))
df1
```

<br>

#### DT Data Table

Install and load DT package

```{r}
library(DT)
datatable(df1)
```

For more info on DT, go to https://rstudio.github.io/DT/

<br>

#### kableExtra

Install and load kableExtra

```{r}
library(kableExtra)

df1[2:4]%>% # columns 2 to 4
  kbl()%>%
  kable_paper(full_width=F, "striped") %>%
  column_spec(2, width="8em", background="yellow") %>%
  pack_rows("Group 1",1,2,label_row_css = "background-color: #666; color: #fff;") %>%
  pack_rows("Group 2",3,5,label_row_css = "background-color: #666; color: #fff;")
```
https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html  

<br>

#### formattable
Install and load formattable  

```{r}
library(formattable)

formattable(df1,  list(`Indicator Name` = formatter("span", style = ~ style(color = "grey", font.weight = "bold")), `y` = color_bar("#FA614B")))
```

<br>
<br>
